Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-20 report/chapter03.Rnw
.

\documentclass[11pt,a4paper,twoside]{book}\usepackage[]{graphicx}\usepackage[]{xcolor}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlsng}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hldef}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\input{header.sty}   % packages, layout and standard macros
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\input{TitelblattMABiostatistics}

\graphicspath{{./figure/}}
\DeclareGraphicsExtensions{.pdf,.png}
\frontmatter
\pagenumbering{roman}
\setcounter{tocdepth}{1}

\chapter*{Abstract}
% TODO: add abstract text.

\chapter*{Acknowledgments}
% TODO: add acknowledgments.

\tableofcontents
\setkeys{Gin}{width=.8\textwidth}

\cleardoublepage
\mainmatter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% LaTeX file for Chapter 01

\chapter{Introduction}\label{ch:intro}

Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty. Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices. A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals. Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background. We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction. All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.

This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence. Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required. The following sections outline our contributions, the research questions we address, and the scope of the work.

\section{Contributions and research questions}

We make three contributions within a single, notation-coherent framework.
\begin{enumerate}
  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
\end{enumerate}

Two questions motivate the empirical study.
\begin{enumerate}
  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines? We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)? We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
\end{enumerate}

\section{Scope and positioning}

We focus on triangular transport maps, transformation random forests, and copulas. Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions. Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used. Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions. Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.

\section{Takeaways and roadmap}

A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail. Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests. Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results. Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% LaTeX file for Chapter 02

\chapter{Methodological Background}\label{ch:background}

We adopt the tutorial's directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a monotone triangular map $S:u\mapsto z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal{N}(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\mathrm{std}}(x))$ with the usual affine Jacobian correction when reporting $\log \pi_X(x)$ (subtract $\sum_{k=1}^K \log \sigma_k$). All Jacobians and partial derivatives involving $S$ are taken in $u$-space unless explicitly stated. This chapter develops the minimal change-of-variables machinery, explains why triangular structure is effective, makes precise the contrast between separable and cross-term triangular maps, and shows how transformation random forests (TRTF) instantiate the same transport with an immediate likelihood identity \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly}.

\section{Change of Variables in Standardized Coordinates}

The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood,
\begin{equation}
  \pi_U(u) = \eta\big(S(u)\big)\, \big|\det \nabla_u S(u)\big|.
  \label{eq:pullback}
\end{equation}
Because we learn $S$ on standardized inputs $u$, Equation~\eqref{eq:pullback} is the form we optimize and evaluate. Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\mathrm{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\mathrm{std}}(x)$. The reference factorizes as $\eta(z)=\prod_{k=1}^K \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions. This linear-in-$K$ structure underpins all evaluation and training in later chapters \citep{ramgraber2025friendly}.

It is useful to record the complementary pushforward statement. If $u\sim \pi_U$ and $z=S(u)$, then $S_{\#}\pi_U=\eta$. The pair $(S^{\sharp}\eta, S_{\#}\pi_U)$ clarifies that the same map both defines the model density in $u$-space via Equation~\eqref{eq:pullback} and certifies exact sampling by inversion. In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in standardized coordinates, and only apply the affine correction when moving results back to the original scale.

\section{Triangular Structure and Its Consequences}

A triangular map decomposes component-wise,
\begin{equation}
  S(u) = \big(S_1(u_1),\,S_2(u_{1:2}),\,\ldots,\,S_K(u_{1:K})\big),
  \label{eq:triangular-map}
\end{equation}
and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence
\begin{equation}
  \det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
  \qquad
  \log \big|\det\nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}).
  \label{eq:triangular-det}
\end{equation}

Three consequences matter in practice. First, likelihoods are efficient: the log determinant in Equation~\eqref{eq:triangular-det} is a sum of one-dimensional terms, so evaluating Equation~\eqref{eq:pullback} scales in $\mathcal{O}(K)$. Second, invertibility is exact: strict monotonicity implies global bijectivity, and inversion reduces to sequential one-dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate \citep{rosenblatt1952remarks,knothe1957contributions}. Third, triangularity yields transparent conditionals: to sample $u_{m:K}\mid u_{1:m-1}=u^{\star}_{1:m-1}$, fix the early coordinates at $u^{\star}$ and invert only the trailing components. Equivalently, the map aligns with the chain-rule factorization $\pi_U(u)=\prod_{k=1}^K \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one-dimensional transformation \citep{ramgraber2025friendly}.

Existence and anisotropy deserve a brief remark. For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe--Rosenblatt rearrangement \citep{rosenblatt1952remarks,knothe1957contributions}. Different orderings induce different maps. While all are valid, their sparsity and approximation difficulty may vary with order. When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling. In high dimension, such sparsity often drives both statistical efficiency and computational cost \citep{ramgraber2025friendly}.

Finally, we keep terminology consistent. A log density is $\log \hat\pi(\cdot)$ at a point. A (test) log likelihood is the dataset average of log densities. Its negative is the negative log likelihood (NLL) in nats; lower is better. These conventions are used uniformly in Chapters~\ref{ch:methods} and \ref{ch:results}.

\section{Separable and Cross-Term Triangular Transports}

A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.

\paragraph{Separable triangular maps.} A separable component decomposes into a context-dependent shift and a univariate monotone shape,
\begin{equation}
  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
  \qquad
  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k) > 0,
  \label{eq:separable-component}
\end{equation}
so the log Jacobian depends only on $u_k$:
\begin{equation}
  \log \partial_{u_k}S_k(u_{1:k}) = \log h_k'(u_k).
  \label{eq:separable-jacobian}
\end{equation}
The context $u_{1:k-1}$ can move the conditional through $g_k$, but it cannot reshape it. Scale, skewness, tail thickness, and modality remain fixed across contexts because the log-Jacobian contribution depends only on $u_k$. Separable transports capture nonlinear location shifts across contexts with a context-invariant shape.

\paragraph{Cross-term triangular maps.} A cross-term component uses an integrated rectifier,
\begin{equation}
  S_k(u_{1:k})=\int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
  \qquad
  \log \partial_{u_k}S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
  \label{eq:cross-component}
\end{equation}
so the log Jacobian, and thus the shape of the conditional, depends on the context. This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.

Two small examples make the contrast concrete. Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$. A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$. A cross-term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$. As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others. Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies. Cross-term training requires careful numerical safeguards; details follow in Chapter~\ref{ch:methods}.

\paragraph{Implementation notes.} The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one-dimensional quadrature and regularization to remain stable. Separable maps often admit linear-in-coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context-invariant conditional shape \citep{ramgraber2025friendly}.

\section{Transformation Forests as Triangular Transport}

Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard, letting predictors act through $h$ while preserving monotonicity \citep{hothorn2018conditional}. A transformation random forest aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ \citep{hothorn2017transformation,hothorn2021transformation}. This induces a triangular transport component via the probability integral transform,
\begin{equation}
  S_k(u_{1:k}) = \Phi^{-1}\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
  \label{eq:trtf-map-theory}
\end{equation}

\paragraph{Per-component likelihood identity.} Differentiating $\Phi\big(S_k(u_{1:k})\big)=\widehat F_k(u_k\mid u_{1:k-1})$ in $u_k$ gives
\begin{equation}
  \widehat\pi_k(u_k\mid u_{1:k-1}) = \phi\big(S_k(u_{1:k})\big)\, \partial_{u_k}S_k(u_{1:k}),
  \label{eq:trtf-likelihood-theory}
\end{equation}
showing that the TRTF conditional density equals the pullback factor one would obtain by parameterizing $S$ directly. Summing over $k$ yields the joint log likelihood from Equation~\eqref{eq:pullback} with the triangular log determinant in Equation~\eqref{eq:triangular-det}. TRTF therefore implements the same transport likelihood once standardized \citep{hothorn2017transformation,hothorn2021transformation,ramgraber2025friendly}.

\paragraph{Additive predictor implies separability.} Under the common additive predictor implementation (TRTF-AP),
\begin{equation}
  \widehat F_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k)+g_k(u_{1:k-1})\big),
  \label{eq:trtf-additive}
\end{equation}
so
\begin{equation}
  S_k(u_{1:k}) = h_k(u_k)+g_k(u_{1:k-1}),
  \qquad
  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k),
  \label{eq:trtf-separable}
\end{equation}
and the induced transport is separable: $\log \partial_{u_k}S_k$ depends only on $u_k$. Monotonicity in $u_k$ is ensured by construction, and the Jacobian term carries no context dependence. TRTF-AP therefore excels when conditional shapes are stable and context acts primarily through location; it is structurally limited when shape varies with predictors \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.

A short note on reporting closes the loop. TRTF evaluates conditional densities in standardized coordinates. Reported log densities on the original scale apply the affine correction $-\sum_{k=1}^K \log \sigma_k$ that accompanies $T_{\mathrm{std}}$, matching the convention set at the beginning of this chapter. With this alignment, comparisons between TRTF, direct triangular maps, and copulas are numerically coherent.

\section{Takeaways}

We learn $S:u\mapsto z$ in standardized coordinates and evaluate $\log \pi_X$ via a simple affine correction. Triangularity turns the log determinant into a sum of one-dimensional terms, guarantees exact inversion, and exposes conditionals through back substitution; these properties make likelihoods linear time in dimension and conditional simulation straightforward \citep{rosenblatt1952remarks,knothe1957contributions,ramgraber2025friendly}. Separable transports move locations but keep shape fixed across contexts; cross-term transports allow context-dependent shape and capture heteroskedastic or multimodal conditionals at the cost of more delicate optimization. TRTF provides a nonparametric route to the same triangular transport with a per-component likelihood identity and the important specialization that TRTF-AP is separable by construction \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}. These foundations justify the objectives and implementations developed in Chapter~\ref{ch:methods} and frame the empirical comparisons that follow.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% LaTeX file for Chapter 03

@@

\textbf{Results.}~Table~\ref{tab:halfmoon-nll} reports joint and conditional NLLs. TTM-X markedly improves over separable and marginal transports, consistent with the need for context-dependent shape. TRTF-AP behaves like a separable transport by construction and underperforms when conditional shape depends on the context, matching the theory in Chapter~\ref{ch:background}. The nonparametric copula performs strongly in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.

\begin{table}[htbp]
  \centering
  \caption{Half-Moon test negative log-likelihoods (nats). Lower is better; $\pm$ indicates standard error.}
  \label{tab:halfmoon-nll}
  \begin{tabular}{lccc}
    \hline
    Model & Mean joint NLL & Conditional NLL 1 & Conditional NLL 2 \\ 
    \hline
    Copula-NP      & $0.87 \pm 0.16$ & $0.76$ & $0.11$ \\ 
    TTM-X          & $1.22 \pm 0.20$ & $0.92$ & $0.29$ \\ 
    TRTF-AP        & $1.83 \pm 0.14$ & $1.25$ & $0.57$ \\ 
    TTM-S          & $1.92 \pm 0.14$ & $1.29$ & $0.64$ \\ 
    TTM-Marginal   & $2.04 \pm 0.12$ & $1.29$ & $0.75$ \\ 
    True marginals & $1.37 \pm 0.11$ & $0.69$ & $0.69$ \\ 
    True joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\ 
    \hline
  \end{tabular}
\end{table}

@@

\textbf{Results.} At very small $n$ the flexible cross-term TTM-X is variance limited, and simpler models (TTM-S, TRTF-AP) can be competitive. As $n$ grows, TTM-X consistently approaches the true joint likelihood, confirming the benefit of cross-term capacity. Conditional NLL decompositions highlight which dimensions drive residual error for separable transports; Table~\ref{tab:autoregressive-nll} illustrates the small-sample regime.

\begin{table}[htbp]
  \centering
  \caption{Four-dimensional autoregressive generator ($n=50$): conditional and joint NLLs (nats). Values are means with standard errors.}
  \label{tab:autoregressive-nll}
  \begin{tabular}{llrrrrrr}
    \hline
    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-S & TTM-X \\ 
    \hline
    1 & Normal      & $1.46 \pm 0.26$ & $1.41 \pm 0.29$ & $1.48 \pm 0.24$ & $1.49 \pm 0.20$ & $1.46 \pm 0.26$ & $1.46 \pm 0.25$ \\ 
    2 & Exponential & $1.55 \pm 0.46$ & $1.38 \pm 0.65$ & $2.54 \pm 0.75$ & $3.30 \pm 0.01$ & $1.75 \pm 0.70$ & $2.58 \pm 0.03$ \\ 
    3 & Beta        & $-0.46 \pm 0.63$ & $-0.63 \pm 1.00$ & $-0.14 \pm 0.34$ & $0.40 \pm 0.17$ & $0.28 \pm 0.70$ & $0.39 \pm 0.23$ \\ 
    4 & Gamma       & $2.21 \pm 1.11$ & $2.07 \pm 0.80$ & $2.22 \pm 1.08$ & $2.78 \pm 0.77$ & $2.97 \pm 1.45$ & $3.00 \pm 1.46$ \\ 
    $K$ & Sum (joint) & $4.75 \pm 1.10$ & $4.23 \pm 1.03$ & $6.10 \pm 1.61$ & $7.97 \pm 0.94$ & $6.47 \pm 1.92$ & $7.43 \pm 1.66$ \\ 
    \hline
  \end{tabular}
\end{table}

@@

\textbf{Interpretation.}~TRTF-AP captures meaningful dependence relative to the independent baseline, but deep flows retain a substantial advantage on this dataset. The gap aligns with the additional flexibility of autoregressive and coupling architectures documented by \citet{papamakarios2017masked} and \citet{dinh2017real}. Conditional PIT diagnostics show that TRTF-AP delivers calibrated marginals by design and partially calibrated conditionals; cross-term transports narrow the remaining deviations at additional computational cost.
\paragraph{Compute.} Triangular models exhibit linear-time evaluation in $K$. Cross-term TTM incurs modest overhead from one-dimensional quadrature and clipping safeguards. TRTF training scales with the number and depth of trees per conditional; evaluation is an ensemble average. Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation thereafter. We additionally report memory footprints where caching affects runtime.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% LaTeX file for Chapter 04

\chapter{Interpretation and Conclusion}\label{ch:conclusion}

This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, highlights methodological insights, and outlines practical guidance and future work.
% Checklist: [x] US spelling [x] Sentence splits [x] Bold list lead-ins

\section{Summary of Findings and Trade-offs}

\subsection{Synthetic benchmarks}

For the Half-Moon dataset ($K=2$), cross-term triangular transport maps (TTM-X) outperform separable transports and transformation forests. This performance confirms the need for context-dependent shape when slices through the data change modality along the curve (Table~\ref{tab:halfmoon-nll}). A low-dimensional nonparametric copula performs competitively in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals. Simpler models (TTM-S, TRTF-AP) remain competitive at small sample sizes because their variance is lower. The four-dimensional autoregressive generator highlights this variance effect (Table~\ref{tab:autoregressive-nll}). TTM-X consistently approaches the true joint likelihood as the sample size grows (Table~\ref{tab:autoregressive-nll}). Conditional NLL decompositions highlight which dimensions drive residual error for separable transports.

\subsection{MINIBOONE}

>>>>

% LaTeX file for Chapter 03
<<'preamble03',include=FALSE>>=
library(knitr)
opts_chunk$set(
    fig.path = 'figure/ch03_fig',
    self.contained = FALSE,
    cache = TRUE
)
@

@@

% Checklist: update NLL table, add runtime table, insert contour figure, reference all elements.
\textbf{Results.}~Table~\ref{tab:halfmoon-nll} summarizes the half-moon test negative log-likelihoods for $n=250$. The true conditional oracle bounds the attainable joint likelihood, while the unconditional copula mixture---formed by weighting each class-conditional fit with the empirical prior---now lies between the oracle and the triangular transports. TTM-Separable and TTM-Marginal remain variance-limited, whereas TRTF closes roughly one third of the gap by exploiting axis-aligned splits. The wall-clock profile in Table~\ref{tab:halfmoon-runtime} shows that these likelihood gains come at the expense of higher training time for TRTF, whereas the copula remains computationally light. Figure~\ref{fig:halfmoon-panels} depicts the resulting log-density contours and illustrates how each estimator captures the upper and lower crescents.

\begin{table}[htbp]
  \centering
  \caption{Half-Moon ($n=250$) test negative log-likelihoods (nats). Lower is better; $\pm$ denotes twice the standard error.}
  \label{tab:halfmoon-nll}
  \begin{tabular}{lccc}
    \hline
    Model & Mean joint NLL & Conditional NLL 1 & Conditional NLL 2 \\ 
    \hline
    True joint        & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\ 
    True conditional  & $0.78 \pm 0.10$ & $0.39$ & $0.39$ \\ 
    Copula mixture    & $1.54 \pm 0.09$ & $0.77$ & $0.77$ \\ 
    TRTF-AP           & $1.71 \pm 0.09$ & $1.23$ & $0.47$ \\ 
    TTM-Separable     & $1.93 \pm 0.08$ & $1.28$ & $0.65$ \\ 
    TTM-Marginal      & $2.02 \pm 0.07$ & $1.28$ & $0.74$ \\ 
    \hline
  \end{tabular}
\end{table}

\begin{table}[htbp]
  \centering
  \caption{Half-Moon ($n=250$) wall-clock timings (seconds) on Apple M1 (8 workers).}
  \label{tab:halfmoon-runtime}
  \begin{tabular}{lcc}
    \hline
    Model & Train time & Test time \\ 
    \hline
    True joint       & $0.000$ & $0.001$ \\ 
    TRTF-AP          & $13.662$ & $1.566$ \\ 
    TTM-Marginal     & $0.003$ & $0.001$ \\ 
    TTM-Separable    & $0.020$ & $0.001$ \\ 
    Copula mixture   & $0.180$ & $0.001$ \\ 
    \hline
  \end{tabular}
\end{table}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.85\textwidth]{halfmoon_panels_seed007_20250919_224351_N250.png}
  \caption{Half-Moon ($n=250$) log-density contours for the true joint, TRTF, TTM variants, and the copula mixture. Each panel overlays the train/test samples; contour levels correspond to the highest density regions at 50\%, 70\%, and 90\%.}
  \label{fig:halfmoon-panels}
\end{figure}

@@

\textbf{Results.} Table~\ref{tab:autoregressive-nll} records the conditional and joint NLLs for the canonical ordering $(1,2,3,4)$ at $n=250$. As expected, the true references align with the ground-truth generator. TRTF-AP tracks the independent exponential component more closely than the marginal map, but once the beta and gamma components enter the context its axis-aligned structure leaves residual multimodality that the triangular transports cannot eliminate. The separable transport mirrors the copula mixture, while the marginal map continues to overfit the exponential tails. These diagnostics motivate the permutation averages in Table~\ref{tab:autoregressive-perm} and Table~\ref{tab:autoregressive-perm-avg}.

\begin{table}[htbp]
  \centering
  \caption{Four-dimensional autoregressive generator ($n=250$, permutation $1,2,3,4$): conditional and joint NLLs (nats). Values are means over test samples.}
  \label{tab:autoregressive-nll}
  \begin{tabular}{llrrrrrr}
    \hline
    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-Separable & Copula mixture \\ 
    \hline
    1 & Normal      & $1.29$ & $1.28$ & $1.28$ & $1.29$ & $1.29$ & $1.30$ \\ 
    2 & Exponential & $1.75$ & $1.49$ & $1.51$ & $2.57$ & $1.88$ & $1.87$ \\ 
    3 & Beta        & $-0.48$ & $-0.79$ & $-0.25$ & $0.28$ & $0.07$ & $0.05$ \\ 
    4 & Gamma       & $2.05$ & $1.83$ & $1.99$ & $2.69$ & $2.41$ & $2.22$ \\ 
    $K$ & Sum (joint) & $4.61$ & $3.80$ & $4.53$ & $6.83$ & $5.66$ & $5.45$ \\ 
    \hline
  \end{tabular}
\end{table}

\textbf{Permutation study.} To check robustness against the ordering assumed by triangular maps, we regenerated all $4!=24$ permutations of $(1,2,3,4)$ at $n=250$. Table~\ref{tab:autoregressive-perm} reports the mean conditional NLL per dimension (and the joint sum) for each estimator averaged over those permutations. The true joint and marginal references remain stable by construction, while TRTF and the separable transport exhibit noticeable shifts as soon as the heavy-tailed components move earlier in the order. The copula mixture averages close to the separable transport, reflecting its independence across conditioning direction. This variability motivates the cross-term map used in the headline results: by enriching each conditional with context-dependent basis functions it removes the dependence on permutation observed for axis-aligned schemes.

\begin{table}[htbp]
  \centering
  \caption{Four-dimensional autoregressive generator ($n=250$): mean test NLL (nats) averaged over all $24$ permutations of $(1,2,3,4)$.}
  \label{tab:autoregressive-perm}
  \begin{tabular}{lrrrrr}
    \hline
    Model & Dim 1 & Dim 2 & Dim 3 & Dim 4 & Sum \\ 
    \hline
    True (joint)       & 1.03 & 0.93 & 0.94 & 0.91 & 3.80 \\ 
    True (marginal)    & 1.22 & 1.13 & 1.15 & 1.11 & 4.61 \\ 
    TRTF-AP            & 1.33 & 1.19 & 1.09 & 1.04 & 4.65 \\ 
    TTM-Marginal       & 1.77 & 1.67 & 1.73 & 1.66 & 6.83 \\ 
    TTM-Separable      & 1.59 & 1.38 & 1.36 & 1.29 & 5.62 \\ 
    Copula mixture     & 1.42 & 1.34 & 1.36 & 1.32 & 5.45 \\ 
    \hline
  \end{tabular}
\end{table}

\begin{table}[htbp]
  \centering
  \caption{Four-dimensional autoregressive generator: permutation-averaged joint test NLL (nats) for $n \in \{25,50,250\}$ (24 permutations per sample size). The $n=100$ column is intentionally left blank because those values are reported elsewhere.}
  \label{tab:autoregressive-perm-avg}
  \begin{tabular}{lrrrr}
    \hline
    Model & $n=25$ & $n=50$ & $n=100$ & $n=250$ \\ 
    \hline
    True (joint)      & 4.35 & 4.23 &  & 3.80 \\ 
    True (marginal)   & 10.50 & 4.75 &  & 4.61 \\ 
    TRTF-AP           & 38.18 & 6.10 &  & 4.64 \\ 
    TTM-Marginal      & 49.36 & 7.43 &  & 6.83 \\ 
    TTM-Separable     & 6829.45 & 6.35 &  & 5.61 \\ 
    Copula mixture    & 9.02 & 6.66 &  & 5.45 \\ 
    \hline
  \end{tabular}
\end{table}

Table~\ref{tab:autoregressive-perm-avg} collects the joint NLL means by parsing the files `main_perm_iter_{n}.txt` and averaging the SUM entries over all $24$ permutations for $n=25$, $n=50$, and $n=250$. The $n=100$ entries are omitted because detailed diagnostics for that setting already appear in the preceding discussion. At $n=25$ the exponential component still drives heavy-tailed behaviour, which explains the enormous averages for the separable and marginal transports; those shrink rapidly once the sample size reaches 50. By $n=250$ the axis-aligned transports settle near $5$--$7$ nats, TRTF drops to $4.6$ nats, and the copula mixture tracks the separable map. The true joint and marginal references remain close to their theoretical values across sample sizes, validating the aggregation procedure.

The synthetic experiments exhaust the main design axes---low-dimensional curvature and higher-dimensional conditional structure. To gauge performance on established density-estimation benchmarks we now turn to the MINIBOONE neutrino dataset and the UCI POWER consumption dataset. Both have become standard testbeds for normalizing flows because they combine moderate dimensionality with non-Gaussian structure; see, for instance, \citet{papamakarios2017masked} and \citet{dinh2017real} for the canonical flow comparisons. Using the same datasets lets us position triangular transports relative to the broader likelihood literature while keeping preprocessing consistent.

@@

\textbf{Interpretation.}~TRTF-AP captures meaningful dependence relative to the independent baseline, but deep flows retain a substantial advantage on this dataset. The gap aligns with the additional flexibility of autoregressive and coupling architectures documented by \citet{papamakarios2017masked} and \citet{dinh2017real}. Conditional PIT diagnostics show that TRTF-AP delivers calibrated marginals by design and partially calibrated conditionals; cross-term transports narrow the remaining deviations at additional computational cost.

\subsection{Real Data: POWER}

\textbf{Setup.} The UCI POWER dataset records household electricity consumption and is routinely used to benchmark likelihood-based generative models alongside MINIBOONE \citep{papamakarios2017masked,dinh2017real}. We follow the standard preprocessing pipeline (log-transform continuous attributes, standardise with train-only statistics, and adopt the canonical train/validation/test split) so that our results remain comparable to published flow numbers.

\textbf{Status.} TRTF experiments on POWER are running at the time of writing; once completed we will tabulate the resulting test log-likelihoods beside the flow baselines to mirror Table~\ref{tab:miniboone-ll}. Preliminary diagnostics indicate similar behaviour to MINIBOONE, with separable transports outperforming independent baselines yet trailing autoregressive flows.

\paragraph{Compute.} Triangular models exhibit linear-time evaluation in $K$. Cross-term TTM incurs modest overhead from one-dimensional quadrature and clipping safeguards. TRTF training scales with the number and depth of trees per conditional; evaluation is an ensemble average. Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation thereafter. We additionally report memory footprints where caching affects runtime.

@@
